\documentclass[12pt, aspectratio=169]{beamer}

\input{../header}
% \usepackage{booktabs}
% \usepackage{tikz}
% \usetikzlibrary{patterns}
% \DeclareMathOperator{\E}{E}
% \DeclareMathOperator{\Var}{Var}
% \DeclareMathOperator{\V}{V}
% \DeclareMathOperator{\C}{C}


\title{Sampling Methods and Common Sampling Distributions}
\author{Md. Aminul Islam Shazid}
\date{}


\begin{document}
    {
		\setbeamertemplate{footline}{}    % NO FOOTLINE FOR THESE TWO FRAMES
		\addtocounter{framenumber}{-2}    % not counting the title page and the outline in frame numbers

		\begin{frame}
			\titlepage
		\end{frame}

		\begin{frame}{Outline}
            \vfill
			\tableofcontents[subsectionstyle=hide]
            \vfill
		\end{frame}
	}


	\section{Introduction}
    \begin{frame}{Sampling}
        \begin{itemize}
            \item In many practical situations it is difficult or impossible to observe every unit in a population

            \item Instead, we select a representative portion or subset of the population and collect information from that subset

            \item This process is called \textbf{sampling}

            \item The selected subset of the population is called a \textbf{sample}
            \end{itemize}

        \begin{block}{Definition}
            Sampling is the process of selecting a representative subset of units from a population in order to study the characteristics of the population
        \end{block}
    \end{frame}


    \begin{frame}{Basic Concepts in Sampling}
        \begin{itemize}
            \item \textbf{Population}

            The entire collection of individuals or units about which information is desired.

            \item \textbf{Sample}

            A subset of the population selected for analysis

            \item \textbf{Sampling Method}

            The procedure used to select the sample from the population
        \end{itemize}
    \end{frame}


    \begin{frame}{Additional Sampling Terminology}
        \begin{itemize}
            \item \textbf{Sampling Unit}

            The basic element or group of elements considered for selection.

            Example: households, students, firms.

            \item \textbf{Sampling Frame}

            A complete list of all sampling units in the population.

            \item \textbf{Sample Unit}

            A sampling unit that is actually selected into the sample.

            \item \textbf{Unit of Inquiry}

            The unit about which information is collected. It may differ from the sampling unit.
        \end{itemize}
    \end{frame}


    \begin{frame}[allowframebreaks]{Example: University Student Survey}

        Suppose a university wants to estimate the \textbf{average daily study time} of its students.

        \begin{itemize}
        \item \textbf{Population}

        All students currently enrolled at the university.

        \item \textbf{Sample}

        A group of 200 students selected from the university to participate in the survey.

        \item \textbf{Sampling Method}

        The procedure used to choose the 200 students, for example simple random sampling from the student list.

        \item \textbf{Sampling Unit}

        Each individual student in the university.

        \item \textbf{Sampling Frame}

        The official list of all enrolled students obtained from the university registrar.

        \item \textbf{Sample Unit}

        The specific 200 students who are selected from the student list.

        \item \textbf{Unit of Inquiry}

        The student about whom information (study hours) is collected.
        \end{itemize}

    \end{frame}


    \begin{frame}{Types of Sampling Methods}
        Sampling methods can broadly be classified into two categories:

        \begin{itemize}
            \item \textbf{Probability Sampling}

            Each unit in the population has a known and non-zero probability of being selected.

            \item \textbf{Non-probability Sampling}

            Selection of units depends on judgement, convenience, or other non-random mechanisms.
        \end{itemize}
    \end{frame}


    \section{Probability Sampling}

    \begin{frame}{Probability Sampling}
        \begin{itemize}
            \item Each element of the population has a known probability of being selected

            \item Randomization is used to select sample units

            \item Sampling bias can be reduced

            \item Statistical inference about the population becomes possible
        \end{itemize}

        \textbf{Common probability sampling methods}

        \begin{itemize}
            \item Simple Random Sampling
            \item Stratified Random Sampling
            \item Systematic Sampling
            \item Cluster Sampling
        \end{itemize}
    \end{frame}


    \subsection{Simple Random Sampling}

    \begin{frame}{Simple Random Sampling (SRS)}
        \begin{block}{Definition}
            A sampling method in which every unit in the population has an equal probability of being selected.
        \end{block}

        \begin{itemize}
            \item If the population size is $N$, the probability of selecting any specific unit is
            \[
                \frac{1}{N}
            \]

            \item Every possible sample of size $n$ has equal probability of being selected
        \end{itemize}
    \end{frame}


    \begin{frame}{Common Selection Methods in SRS}
        Selection method means how items are selected into the sample.
        \begin{itemize}
            \item Lottery method
            \item Random number table
            \item Computer generated random numbers
        \end{itemize}
    \end{frame}


    \subsection{Stratified Random Sampling}

    \begin{frame}{Stratified Random Sampling}
        \begin{block}{Definition}
            In stratified sampling the population is divided into several non-overlapping groups called \textbf{strata}, and a random sample is selected from each stratum.
        \end{block}

        \begin{itemize}
            \item Strata are formed using known characteristics

            \item Units within each stratum are relatively homogeneous
            
            \item Units across different strata ar heterogeneous

            \item Sampling is performed independently within each stratum
        \end{itemize}
    \end{frame}


    \begin{frame}{Procedure of Stratified Random Sampling}
        \begin{enumerate}
            \item Divide the population into $L$ strata

            \item Determine total sample size $n$

            \item Allocate sample sizes $n_1, n_2, \dots, n_L$ to each stratum

            \item Select a random sample from each stratum
        \end{enumerate}

        \begin{align*}
            \text{Population sizes of each strata: } & N_1 + N_2 + \cdots + N_L = N \\
            \text{Sample sizes of each strata: } & n_1 + n_2 + \cdots + n_L = n
        \end{align*}
    \end{frame}

    
    \begin{frame}{Ways to Allocate Sample Size to Each Stratum}
        In stratified sampling there are a few ways to allocate the total sample size \(n\) among the strata:
        \begin{itemize}
            \item Equal sample size in each strata: \(n_1 = n_2 = ... = n_L\)
            \item Sample size proportional to strata population size: \(n_i = (n/N) \times N_i\). It is also known as probaility proportional to size
            \item Sample size proportional to strata variance: \(n_i = (n/N) \times V_i\), where \(V_i\) is the variance of the strata \(i\) (this is usually not known in advance, guess work or prior research is used to approximate \(V_i\)). It is also known as probability proportional to variance
        \end{itemize}
    \end{frame}


    \subsection{Systematic Sampling}

    \begin{frame}{Systematic Sampling}
        \begin{block}{Idea}
            Sample units are selected at regular intervals from an ordered list of the population.
        \end{block}

        \begin{itemize}
            \item Population size: $N$

            \item Sample size: $n$

            \item Sampling interval:
            \[
                k = \frac{N}{n}
            \]
        \end{itemize}
    \end{frame}


    \begin{frame}{Procedur of Systematic Sampling}
        \begin{enumerate}
            \item Number the population units from $1$ to $N$

            \item Compute the sampling interval $k = N/n$

            \item Randomly select the first unit from $1,2,\dots,k$

            \item Select subsequent units as
            \[
                r, r+k, r+2k, \dots
            \]
        \end{enumerate}
    \end{frame}


    \subsection{Cluster Sampling}

    \begin{frame}{Cluster Sampling}
        \begin{block}{Definition}
            Cluster sampling is a method in which the population is divided into groups called clusters and a random sample of clusters is selected.
        \end{block}

        \begin{itemize}
            \item Each cluster contains multiple population elements

            \item All elements in selected clusters may be included in the sample

            \item Often used when a complete sampling frame is not available
        \end{itemize}
    \end{frame}


    \section{Non-probability Sampling}

    \begin{frame}{Non-probability Sampling}
        \begin{itemize}
            \item In non-probability sampling, the probability of selection is not known

            \item Selection depends on judgement, accessibility, or convenience

            \item These methods are easier to implement but may introduce bias
        \end{itemize}

        \textbf{Common types}

        \begin{itemize}
            \item Convenience sampling
            \item Purposive sampling
            \item Judgement sampling
            \item Quota sampling
            \item Snowball sampling
        \end{itemize}
    \end{frame}


    \begin{frame}[allowframebreaks]{Examples of Non-probability Sampling}
        \begin{itemize}
            \item \textbf{Convenience Sampling}: Selecting individuals who are easiest to reach.
            
            Example: A researcher studying smartphone usage asks students in his own class to complete a survey because they are easy to reach.
            
            \item \textbf{Judgment Sampling}: Researcher selects units based on judgment. 
            
            Example: An expert selects a group of companies that he believes are representative of the technology sector.

            \item \textbf{Purposive Sampling}: Selecting individuals based on specific characteristics.
            
            Example: A study on experienced surgeons selects only doctors who have performed more than 500 surgeries.

            \item \textbf{Quota Sampling}: Selecting units to satisfy predefined group proportions.
            
            Example: A researcher wants a sample of 100 people consisting of 50 males and 50 females. Interviewers select individuals until each quota is filled.

            \item \textbf{Snowball Sampling}: Existing participants recruit additional participants.
            
            Example: A researcher studying illegal drug use interviews a few participants and asks them to refer other users they know.
        \end{itemize}
    \end{frame}


    \begin{frame}{Summary}
        \begin{itemize}
            \item Sampling allows us to study large populations efficiently

            \item Sampling methods are broadly classified as probability and non-probability methods

            \item Probability sampling supports statistical inference

            \item Common probability methods include simple random, stratified, systematic, and cluster sampling
        \end{itemize}
    \end{frame}


    \section{Law of Large Number and Central~Limit~Theorem}

    \begin{frame}{Law of Large Numbers}
        \begin{block}{Idea}
            As the sample size increases, the sample mean tends to get closer to the population mean.
        \end{block}

        \begin{itemize}
            \item Let $\mu$ denote the population mean

            \item Let $\bar{X}$ denote the sample mean

            \item When the sample size $n$ becomes large,
            \[
            \bar{X} \rightarrow \mu
            \]

            \item Thus, larger samples generally provide more reliable estimates of the population mean
        \end{itemize}
    \end{frame}


    \begin{frame}{Law of Large Numbers (Interpretation)}
        \begin{itemize}
        \item Suppose we repeatedly draw samples from the same population

        \item For small samples, the sample mean may vary substantially from the population mean

        \item As the sample size increases, the variability of the sample mean decreases

        \item Therefore, the sample mean becomes a better approximation of the population mean
        \end{itemize}
    \end{frame}


    \begin{frame}{Central Limit Theorem}
        \begin{block}{Statement}
            For a sufficiently large sample size, the sampling distribution of the sample mean is approximately normal, regardless of the shape of the population distribution.
        \end{block}

        If the population mean is $\mu$ and population standard deviation is $\sigma$, then

        \[
            \bar{X} \sim N\left(\mu,\frac{\sigma^2}{n}\right)
        \]

        for sufficiently large $n$.
    \end{frame}


    \begin{frame}{Implications of the Central Limit Theorem}
        \begin{itemize}
            \item The distribution of the sample mean approaches a normal distribution as the sample size increases

            \item This holds even when the population distribution is not normal

            \item The standard deviation of the sampling distribution of the sample mean is
            \[
                \frac{\sigma}{\sqrt{n}}
            \]
            which is called the \textbf{standard error of the mean}

            \item When sample size increases the standard error decreases
        \end{itemize}
    \end{frame}


    \section{Sampling Distributions}
    
    \begin{frame}{Sampling Distributions}
        \begin{itemize}
            \item The value of a statistic varies across different samples
            \item A \textbf{sampling distribution} is the theoretical probability distribution of a statistic computed from repeated random samples of a population
            \item Common statistics include the sample mean, sample variance, and ratios of variances
            \item Sampling distributions describe how these statistics vary from sample to sample
            \item They form the theoretical basis for statistical inference and hypothesis testing
        \end{itemize}
    \end{frame}


    \begin{frame}{Why Sampling Distributions Matter}
        \begin{itemize}
            \item Population parameters are usually unknown
            \item Instead, we compute statistics from samples. A statistic is used to estimate a parameter
            \item Sampling distributions allow us to:
            \begin{itemize}
                \item quantify sampling variability,
                \item calculate probabilities for observed statistics,
                \item construct confidence intervals,
                \item perform hypothesis tests.
            \end{itemize}
        \end{itemize}
    \end{frame}


    \begin{frame}{Degrees of Freedom}
        \begin{itemize}
            \item \textbf{Degrees of freedom (DoF)} refer to the number of independent pieces of information available when estimating a parameter

            \item In many statistical calculations, some information is used to estimate parameters. This reduces the number of values that are free to vary

            \item Example: suppose we have a sample of size \(n\) and we compute the sample mean \(\bar{X}\)

            \item Once the mean is fixed, only \(n-1\) observations can vary freely. The final observation is determined by the requirement that the sample mean remains unchanged

            \item Therefore, the sample variance uses \(n-1\) degrees of freedom

            \item Many sampling distributions depend on degrees of freedom, including Chi-square distribution, Student's \(t\) distribution, \(F\) distribution
        \end{itemize}
    \end{frame}


    \subsection{Chi-Square Distribution}

    \begin{frame}{Chi-Square \((\chi^2)\) Distribution}
        \begin{itemize}
            \item If \(Z_1, Z_2, \dots, Z_k\) are independent standard normal variables, then
            \[
                \chi^2 = Z_1^2 + Z_2^2 + \dots + Z_k^2
            \]
            follows a \textbf{Chi-Square distribution} with \(k\) degrees of freedom
            \item The distribution depends on the parameter called \textbf{degrees of freedom}
            \item It is defined only for positive values
            \item The distribution is right-skewed, especially for small degrees of freedom
        \end{itemize}
    \end{frame}


    \begin{frame}{Use of Chi-Square Distribution}
        \begin{itemize}
            \item Testing hypotheses about population variance
            \item Goodness-of-fit tests
            \item Tests of independence in contingency tables
            \item Many statistical procedures rely on the Chi-square distribution of the sample variance
        \end{itemize}
    \end{frame}


    \subsection{Student's \(t\) Distribution}

    \begin{frame}{Student's \(t\) Distribution}
        \begin{itemize}
            \item Let \(\bar{X}\) be the sample mean of the random variable \(X\). Then \(\bar{X}\) is also a random variable because the sample mean varies across different samples
            \item The statistic
            \[
                t = \frac{\bar{X} - \mu}{s/\sqrt{n}}
            \]
            follows a \(t\)-distribution with \(n-1\) degrees of freedom, where:
            \begin{itemize}
                \item \(\mu\) is the population mean
                \item \(s\) is the sample standard deviation of the observed data
                \item \(n\) is the sample size
            \end{itemize}
            \item The distribution is symmetric around zero
        \end{itemize}
    \end{frame}


    \begin{frame}{Use of \(t\) Distribution}
        \begin{itemize}
            \item Testing hypotheses about a population mean when variance is unknown
            \item Constructing confidence intervals for the mean
        \end{itemize}
    \end{frame}


    \subsection{\(F\) Distribution}

    \begin{frame}{\(F\) Distribution}
        \begin{itemize}
            \item The \textbf{\(F\) distribution} arises from the ratio of two independent Chi-square variables
            \item If
            \[
                F = \frac{(U_1 / d_1)}{(U_2 / d_2)}
            \]
            where \(U_1 \sim \chi^2(d_1)\) and \(U_2 \sim \chi^2(d_2)\), then \(F\) follows an \(F(d_1,d_2)\) distribution
            \item The distribution depends on two degrees of freedom parameters
            \item It takes only positive values and is right-skewed
        \end{itemize}
    \end{frame}


    \begin{frame}{Use of \(F\) Distribution}
        \begin{itemize}
            \item Comparing two population variances
            \item Analysis of variance (ANOVA)
            \item Regression model significance testing
        \end{itemize}
    \end{frame}


    \section*{Thank you.}

\end{document}