\documentclass[12pt, aspectratio=169]{beamer}

\input{../header}
% \usepackage{booktabs}
% \usepackage{tikz}
% \usetikzlibrary{patterns}
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\V}{V}
\DeclareMathOperator{\C}{C}


\title{Random Variable and Probabiilty Distribution}
\author{Md. Aminul Islam Shazid}
\date{}


\begin{document}
    {
		\setbeamertemplate{footline}{}    % NO FOOTLINE FOR THESE TWO FRAMES
		\addtocounter{framenumber}{-2}    % not counting the title page and the outline in frame numbers

		\begin{frame}
			\titlepage
		\end{frame}

		\begin{frame}{Outline}
            \vfill
			\tableofcontents[subsectionstyle=hide]
            \vfill
		\end{frame}
	}

	\section{Random Variable}

    \begin{frame}{Introduction}
        A variable which takes numerical values resulting from random experiments, is called a random variable (RV).

		\begin{itemize}
			\item There is a probability associated with each possible values
			\item Random variables are denoted by capital letters such as $X, Y, Z$ etc.
			\item Possible values are denoted by small letters such as $x, y, z$ etc.
			\item Example:
			\begin{itemize}
				\item Height of students
				\item Number of heads when tossing a coin three times
			\end{itemize}
		\end{itemize}
    \end{frame}


	\begin{frame}{Types of Random Variable}
		\begin{itemize}
			\item \textbf{Discrete random variable}: A random variable defined over a discrete sample space
			\begin{itemize}
				\item Number of students in a class, sample space, $S = \{0, 1, 2, ..., \infty\}$
				\item Number of correct answers in among $50$ questions, $S = \{0, 1, 2, ..., 50\}$
			\end{itemize}
			\item \textbf{Continuous random variable}: A random variable defined over a continuous sample space
			\begin{itemize}
				\item Monthly income, $S = \{X: 0 \leq x < \infty\}$
				\item Monthly profit, $S = \{X: -\infty < x < \infty\}$
			\end{itemize}
		\end{itemize}
	\end{frame}


	\section{Probability Distribution}

	\begin{frame}{Probability Distribution}
		\begin{itemize}
			\item Probability distribution means the distribution of the probabilities among the different values of a random variable
			\item For example, when tossing a coin twice, the sample space, $S = \{HH, HT, TH, TT\}$
			\item Let an RV, $X = $ number of heads in two coin tosses, then $X$ can take the values: $0, 1, 2$
			\item Then the probability of each value of $X$:
			\begin{center}
				\begin{tabular}{c|c}
					\hline
					$x$ & $P(X = x)$ \\
					\hline
					$0$ & $1/4$ \\
					$1$ & $2/4$ \\
					$2$ & $1/4$ \\
					\hline
				\end{tabular} \label{cointosstable}
			\end{center}
		\end{itemize}
	\end{frame}


	\begin{frame}{Types of Probability Distributions}
		Depending on the type of variable, distributions are of two types:
		\begin{itemize}
			\item \textbf{Discrete probability distribution}: probability distribution of a discrete random variable
			\item \textbf{Continuous probability distribution}: probability distribution of a continuous random variable
		\end{itemize}
	\end{frame}


	\begin{frame}{Discrete probability distribution}
		\begin{itemize}
			\item The probability distribution of a discrete random variable is a table, graph, formula, or other device used to specify all possible values of a discrete random variable along with their respective probabilities
			\item If we let the discrete probability distribution be represented by the function $p(x)$, then $p(x) = P(X = x)$ is the probability of the discrete random variable $X$ to assume a value $x$
			\item $p(x)$ is called a probability mass function (PMF)
		\end{itemize}
	\end{frame}


	\begin{frame}{Probability Mass Function (PMF)}
		A probability mass function gives the probability of each of the value of a discrete random variable.

		\vspace{0.5em}

		A function, $p(x)$, of a discrete random variable $X$ will be called a PMF, if and only if all the following conditions are satisfied:
		\begin{enumerate}
			\item $p(x) \ge 0; \, \forall x$
			\item $\displaystyle \sum_{x} p(x) = 1$
			\item $P(X = a) = p(a)$
		\end{enumerate}
	\end{frame}


	\begin{frame}{Example: PMF}
		In page \ref{cointosstable}, the probability mass function of the variable $X$ representing the number of heads when a coin is tossed twice is given. It is an example of a PMF.

		\begin{center}
				\begin{tabular}{c|c}
					\hline
					$x$ & $P(X = x)$ \\
					\hline
					$0$ & $1/4$ \\
					$1$ & $2/4$ \\
					$2$ & $1/4$ \\
					\hline
				\end{tabular}
			\end{center}
	\end{frame}


	\begin{frame}{Probabiilty Desity Function (PDF)}
		The probability distribution of a continuous random variable is called continuous probability distribution. The probabilities are given by the probability density function (PDF) of that distribution.

		\vspace{0.5em}

		A function $f(x)$ of a continuous random variable $X$ is called the PDF of $X$ if and only if the following conditions are satisfied:
		\begin{enumerate}
			\item $f(x) \ge 0; \, \forall x$
			\item $\int_x f(x) dx = 1$
			\item $P(a \le X \le b) = \int_a^b f(x) dx$
		\end{enumerate}
		Here, $\int_x f(x) dx$ means integration of $f(x)$ over the whole sample space of $X$. Sometimes, it's also written as $\int_{-\infty}^{\infty} f(x) dx$.
	\end{frame}


	\begin{frame}{PDF: Example}
		Suppose $X$ is a continuous random variable whose density function is given by:
		\[
			f(x) = \begin{cases}
					C(4x - 2x^2) \quad & \text{if}\, 0 < x < 2 \\
					0 & \text{otherwise}
				\end{cases}
		\]
		\begin{enumerate}
			\item What is the value of $C$?
			\item Find $P(X>1)$.
		\end{enumerate}
	\end{frame}


	\begin{frame}{PDF: Example Solution}
		\begin{enumerate}
			\item \begin{align*}
				& \int_x f(x) dx = 1 \\
				\Rightarrow & \int_0^2 C(4x - 2x^2) dx = 1 \\
				\Rightarrow &\,  C\big[4x^2/2 - 2x^3/3\big]_0^2 = 1 \\
				\Rightarrow &\, C = 3/8
			\end{align*}

			\item \[
				P(X>1) = \int_1^2 \frac{3}{8} (4x - 2x^2) = 1/2
			\]
		\end{enumerate}
	\end{frame}
	

	\begin{frame}{Cumulative Distribution Function (CDF)}
		The cumulative distribution function (CDF) gives the probability that a random variable $X$ is lower than some value $x$. Sometimes, it is also called ``distribution function''.

		\begin{itemize}
			\item For discrete random variable, it is given by  $P(X<x) = \displaystyle \sum_{k=-\infty}^{x} p(k)$
			\item For continuous case, it is denoted by $F(x)$
			\[
				F(x) = P(X<x) = \int_{-\infty}^{x} f(x) dx
			\]
			\item It is a non-decreasing function
		\end{itemize}

		Here, summing/integrating from $-\infty$ means that the lower limit of the sum/integral is the smallest value in the domain of the random variable $X$.
	\end{frame}


	\begin{frame}{CDF: Example 1}
		In page \ref{cointosstable}, the probability mass function of the variable $X$ representing the number of heads when a coin is tossed twice is given.

		Below, its CDF is also given:

		\begin{center}
				\begin{tabular}{c|c|c}
					\hline
					$x$ & $P(X = x)$ & $P(X \le x)$\\
					\hline
					$0$ & $1/4$ & $1/4$\\
					$1$ & $2/4$ & $3/4$\\
					$2$ & $1/4$ & $1$\\
					\hline
				\end{tabular}
			\end{center}
	\end{frame}


	\begin{frame}{CDF: Example 2}
		Suppose $X$ is a continuous random variable whose density function is given by: \label{pdf}
		\[
			f(x) = \begin{cases}
					\frac{3}{8}(4x - 2x^2) \quad & \text{if}\, 0 < x < 2 \\
					0 & \text{otherwise}
				\end{cases}
		\]
		Find the CDF of $X$:

		\begin{align*}
			F(X) &= \int_{0}^{x} f(x) dx \\
			&=  \int_{0}^{x} \frac{3}{8}(4x - 2x^2) dx \\
			&= \frac{3}{8} \big[(4x^2/2 - 2x^3/3) \big]_0^x
			= \frac{3}{8} \big[(2x^2 - 2x^3/3) - 0 \big]
			= \frac{3}{8} (2x^2 - 2x^3/3)
		\end{align*}
	\end{frame}


	\section{Mathematical Expectation and Variance}


	\begin{frame}{Mathematical Expectation}
		Expectation of an RV, $X$ is the weighted average of the possible values of $X$. The weight of each value of $X$ is its corresponding probability.

		\begin{itemize}
			\item When $X$ is discrete: \(\E(X) = \displaystyle \sum_{x} x\cdot p(x)\)
			\item When continuous: \(\E(X) = \displaystyle \int_{x} x\cdot f(x) dx\)
			\item The sum/integral is over all values of \(X\)
			\item It is often denoted by \(\mu\), that is: \(\E(X) = \mu\)
			\item It is also known as expected value or population mean
		\end{itemize}
	\end{frame}


	\begin{frame}{Variance of a Random Variable}
		The variance of an RV, \(X\) measures the dispersion in the values of \(X\). It is denoted by \(\mathrm{Var}(X)\) or \(\mathrm{V}(X)\).

		\begin{align*}
			\V(X) &= \E[(X-\E(X))^2] \\
			&= \E[X^2 - 2X \cdot \E(X) + [\E(X)]^2] \\
			&= \E(X^2) - 2E(X)\cdot \E(X) + [\E(X)]^2 \\
			&= \E(X^2) - 2[\E(X)]^2 + [\E(X)]^2 \\
			&= \E(X^2) - [\E(X)]^2
		\end{align*}

		Here, \(\E(X^2)\) is given by \(\displaystyle \sum_x x^2 \cdot p(x)\) or \(\displaystyle \int_x x^2 \cdot f(x) \, dx\) for discrete and continuous case respectively.
	\end{frame}


	\begin{frame}{Expectation and Variance: Example 1}
		In page \ref{cointosstable}, the PMF of the random variable \(X\) representing the number of heads in two throws of a fair coin is given.
		
		\(X\) can take the values $0$, $1$ and $2$ with probabilities $1/4$, $1/2$ and $1/4$ respectively.

		The expected value of $X$ is:
		\[
			\E(X) = 0 \cdot \frac{1}{4} + 1 \cdot \frac{1}{2} + 2 \cdot \frac{1}{4} = 0 + \frac{1}{2} + \frac{1}{2} = 1
		\]

		The variance of $X$ is:
		\begin{align*}
			\V(X) &= \E(X^2) - [\E(X)]^2 \\
			&=  \big[0^2 \cdot \frac{1}{4} + 1^2 \cdot \frac{1}{2} + 2^2 \cdot \frac{1}{4}\big] - 1^2 \\
			&= \big[0 + \frac{1}{2} + 1 \big] - 1 = \frac{1}{2}
		\end{align*}
	\end{frame}


	\begin{frame}{Expectation and Variance: Example 2}
		Find the expected value and the variance of the RV $X$ whose PDF is given in page \ref{pdf}.
		\begin{align*}
			\E(X) &= \int_{0}^{2} x \cdot \frac{3}{8}(4x - 2x^2) dx \\
			&= \frac{3}{8} \int_{0}^{2} (4x^2 - 2x^3) dx \\
			&= \frac{3}{8} \big[4x^3/3 - 2x^4/4 \big]_0^2 \\
			&= \frac{3}{8} \big[4 \times 2^3/3 - 2\times 2^4/4 - 0 \big] = 1
		\end{align*}
		Finding the variance is left as an exercise.
	\end{frame}


	\begin{frame}{Properties of Expecation and Variance}
		Let $a$, $b$ and $C$ be some constants, and $X$ and $Y$ be two independent random variables.
		\begin{columns}
			\begin{column}{0.5\textwidth}
				\begin{enumerate}
					\item \(\E(c) = c\)
					\item \(\E(cX) = c \E(X)\)
					\item \(\E(c + X) = c + \E(X)\)
					\item \(\E(aX + bY) = a \E(X) + b \E(Y)\)
				\end{enumerate}
			\end{column}
			
			\begin{column}{0.5\textwidth}
				\begin{enumerate}
					\item \(\V(c) = 0\)
					\item \(\V(cX) = c^2 \V(X)\)
					\item \(\V(c + X) = c + \V(X)\)
					\item \(\V(aX + bY) = a^2 \V(X) + b^2 \V(Y)\)
				\end{enumerate}
			\end{column}
		\end{columns}
		
		\vspace{1em}

		If $X$ and $Y$ are dependent: \(\V(aX + bY) = a^2 \V(X) + b^2 \V(Y) + 2ab \operatorname{CoV}(a,b)\).

		\vspace{0.5em}
		
		Here, $\operatorname{CoV}(a,b)$ is the covariance of $X$ and $Y$. More on covariance in a future lecture.
	\end{frame}


	\begin{frame}{Properties of Expecation and Variance: Example 1}
		Suppose $X$ is a random variable with \(\E(X) = 5\) and \(\V(X)=2\).

		\begin{itemize}
			\item \(\E(2X) = 2 \E(X) = 2 \times 5 = 10\)
			\item \(\V(2X) = 2^2 \V(X) = 4 \times 2 = 8\)
			\item \(\E(2X - 3) = 2 \E(X) - 3 = 2 \times 5 - 3= 10 - 3 = 7\)
			\item \(\V(2X - 3) = 2^2 \V(X) - 3 = 4 \times 2 - 3 = 8 - 3 = 5\)
		\end{itemize}
	\end{frame}	


	% \begin{frame}{Properties of Expecation and Variance: Exercise}
		
	% \end{frame}


	\section{Some Common Discrete Probability Distributions}


	\begin{frame}{Bernoulli Process}
		When a random process or experiment, called a trial, can result in only one of two mutually exclusive outcomes, such as dead or alive, sick or well, full-term or premature, the trial is called a Bernoulli trial.

		\begin{itemize}
			\item Each trial results in one of two possible, mutually exclusive, outcomes. One of the possible outcomes is denoted (arbitrarily) as a success, and the other is denoted a failure
			\item The probability of a success, denoted by $p$, remains constant from trial to trial. The probability of a failure, $1 - p$, is denoted by $q$
			\item The trials are independent; that is, the outcome of any particular trial is not affected by the outcome of any other trial
		\end{itemize}
	\end{frame}


	\begin{frame}{Bernoulli Distribution}
		Let $X$ be the number of successe in a single Bernoulli trial. Then $X$ follows a Bernoulli distribution. It is written as:
		\[X \sim \mathrm{Bernoulli}(p).\]

		The probability mass function (PMF) of $X$ is given as:
		\[
			p(x) = p^x (1-p)^{1-x} \, ; \quad x = 0, 1
		\]

		\begin{itemize}
			\item Expected value, \(\E(X) = 1 \times p + 0 \times (1-p) = 1\)
			\item Variance: \begin{align*}
				\V(X) = \E(X^2) - [\E(X)]^2 &= (1^2 \times p + 0^2 \times (1-p)) - p^2 \\
				 &= p^2 - p = p(1-p)
			\end{align*}
		\end{itemize}
	\end{frame}


	\begin{frame}{Binomial Distribution}
		Let $X$ be the number of successes in $n$ Bernoulli trials, where the probability of success is $p$. Then the random variable $X$ follows a binomial distribution. It is written as:
		\[X \sim \mathrm{binom}(n, p).\]

		The probability mass function (PMF) of $X$ is given as:
		\[
			p(x) = ^n\C_x \, p^x (1-p)^{n-x} \, ; \quad x = 0, 1, 2, 3...
		\]

		\begin{itemize}
			\item \(\E(X) = np\)
			\item \(\V(X) = np(1-p)\)
		\end{itemize}
	\end{frame}


	\begin{frame}{Examples of Binomial Variables}
		\begin{itemize}
			\item Number of defective products among a batch of products
			\item Number of patients carrying a ceratain pathogen among a group of patients
			\item Number of heads when a coin is tossed several times
		\end{itemize}
	\end{frame}


	\begin{frame}{Binomial Distribution: Example}
		There are 3 multiple choice questions in a MCQ test. Each MCQ consists of four
possible choices and only one of them is correct. If an examinee answers those
MCQ randomly (without knowing the correct answers)
		\begin{enumerate}
			\item What is the probability that exactly any two of the answers will be correct?
			\item What is the probability that at least two of the answers will be correct?
			\item What is the probability that at most two of the answers will be correct?
			\item What will be the average or expected number of correct answers?
			\item Also, find the standard deviation of number of correct answers.
		\end{enumerate}
	\end{frame}


	\begin{frame}{Binomial Distribution: Example Solution}
		Each MCQ has $4$ options. So, when answering randomly, probability of correct answer, $p = 1/4 = 0.25$.
		Let the random variable $X$ denote the number of correct answers. Then, \(X \sim \mathrm{binomial}(3, 0.25)\).

		\begin{enumerate}
			\item Probability of two correct answers, \(P(X=2) = ^3\C_2 \, 0.25^2 \, (1-0.25)^{3-2}\)
			\item Probability of \emph{at least} two correct answers \(=P(X=2) + P(X=3)\)
			\item Probability of \emph{at most} two correct answers \(=P(X=0) + P(X=1) + P(X=2)\)
			\item Expected number of correct answers, \(\E(X) = 3 \times 0.25 = 0.75\)
			\item Variance of number of correct answers, \(\V(X) = 3 \times 0.25 \times (1-0.25)\)
		\end{enumerate}

		The calculations for question 2 and 3, are left as exercises.
	\end{frame}


	\begin{frame}{Poisson Distribution}
		A Poisson random variable is a discrete random variable that counts the number of times a certain event occurs within fixed interval of time, or area, or volume or any other unit.

		\vspace{0.5em}
		
		The Poisson distribution describes the probability of these counts.

		\vspace{0.5em}

		Let \(X ~ \sim \mathrm{Poisson}(\lambda)\), then the PMF of $X$ is given by:
		\[
			P(X=x) = \frac{e^{-\lambda} \lambda^x}{x!} \,; \quad \lambda > 0; \, \quad x=0, 1, 2, 3,...
		\]

		\begin{itemize}
			\item $\lambda$ is the parameter of the distribution
			\item \(\E(X)=\lambda\)
			\item \(\V(X)=\lambda\)
		\end{itemize}
	\end{frame}


	\begin{frame}{Assumptions of a Poisson Process}
		\begin{itemize}
			\item Events occur independently of each other
			\item The probability of more than one event in a very small interval is negligible
			\item The average rate of occurrence $\lambda$ is constant over the interval
			\item The probability of an event occurring is proportional to the length of the interval
		\end{itemize}
	\end{frame}


	\begin{frame}{Examples of Poisson Random Variables}
		\begin{itemize}
			\item Number of phone calls received by a call center in one minute
			\item Number of typing errors on a page of text
			\item Number of accidents at an intersection in a day
			\item Number of emails arriving in an hour
		\end{itemize}
	\end{frame}


	\begin{frame}{Poisson Distribution: Example}
		The average number of errors on a page of a certain magazine is 0.2. What is the probability that:
		\begin{enumerate}
			\item a randomly selected page contains $0$ (zero) error?
			\item a randomly selected page contains $2$ or more errors?
			\item two consecutive pages have $3$ errors?
			\item What is the average error per page?
			\item Also, find standard deviation of the number of errors in a page.
		\end{enumerate}
	\end{frame}


	\begin{frame}{Poisson Distribution: Example Solution}
		Here, error rate, \(\lambda = 0.2\).
		\begin{enumerate}
			\item Probability that a randomly selected page contains $0$ errors is 
			\[
				P(X=0) = \frac{e^{-0.2} \, 0.2^0}{0!}
			\]
			\item Probability that a randomly selected page contains $2$ or more errors is
			\begin{align*}
				P(X \geqslant 2) &= 1 - P(X < 2) \\
				&= 1 - [P(X=0) + P(X=1)]
			\end{align*}
		\end{enumerate}
		The above calculations are left as an exercise.
	\end{frame}


	\begin{frame}{Poisson Distribution: Example Solution (cont.)}
		
		\begin{enumerate}
			\setcounter{enumi}{2}
			\item Probability that two consecutive pages have $3$ errors:
			
			\vspace{0.25em}
			
			The error rate for two pages is, \(\lambda_{new} = 0.2 \times 2 = 0.4\)/

			\vspace{0.25em}

			Therefore, the desired probability is:
			\[
				\frac{e^{-0.4} \, 0.4^3}{3!}.
			\]

			\item Average error per page is \(\E(X) = \lambda = 0.2\).
			\item Standard deviation of number of errors in a page, \(\operatorname{SD}(X) = \sqrt{\V(X)} = \sqrt{0.2}\).
		\end{enumerate}
	\end{frame}


	\begin{frame}{Geometric Distribution}
		
	\end{frame}


	\section{Some Common Continuous Probability Distributions}


	\begin{frame}{Uniform Distribution}
		
	\end{frame}


	\begin{frame}{Normal Distribution}
		
	\end{frame}


	\begin{frame}{Exponential Distribution}
		
	\end{frame}

	
	\section{Approximating Probabilities using the Normal Distribution}


    \section*{Thank you.}

\end{document}